ICT and Future Productivity:\\Evidence and Theory of a GPT\thanks{Correspondence: Department of Economics, Boston College, 140 Commonwealth Avenue, Chestnut Hill, MA 02467. Email: brianti@bc.edu (Marco Brianti) and gati@bc.edu (Laura G\'ati).}

### 2018-07-30 Latex/Draft_Summer2018/Summer2018.tex
read through the whole paper
ready to discuss some stuff

Secondly, a long line of literature going back to the early 1990s postulates that ICT may be a general-purpose technology.\footnote{\cite{basu2003case}, \cite{basu2007information}, \cite{fernald2017disappointing} and \cite{oliner2000resurgence} and are some examples.} Following the classical definition of \cite{bresnahan1992general}, GPTs are technologies that alter the process of economy-wide production, and thus affect the productivity of the entire economy.

The general-purpose nature of ICT capital works through its own diffusion because information sharing improves only when new communication technologies are widely spread over the population.

@@

Consider an economy inhabited by a representative agent who maximizes the expected values of his lifetime utility as given by
\begin{eqnarray}\label{equation:utilFunc}
\E \Bigg[  \sum_{t=0}^{\infty} \beta^t U(c_t,l_t)  \Bigg]
\end{eqnarray}
with 
%\begin{eqnarray*}
%U(c_t,l_t) = \log(c_t) - \frac{\chi}{2} l_t^{2}
%\end{eqnarray*}
\begin{eqnarray*}
U(c_t,l_t) = \log(c_t) - \frac{1}{1+\frac{1}{\chi}} l_t^{1+\frac{1}{\chi}}
\end{eqnarray*}
where $c_t$ and $l_t$ represent consumption and labor at time $t$.

@@

In this section, we estimate the model's key parameters governing the effect of ICT investment on future TFP by imposing theoretical impulse responses of an ICT shock to be as close as possible to the empirical ones. To enhance movement in ICT investment, we use a shock to the ICT-specific technological component $\zeta_t$. Using empirical impulse responses of TFP, ICT investment, consumption, and relative prices, we estimate three parameters: the variance of an ICT technology shock, $\sigma_{\iota}^2$, the persistence of the same shock, $\rho_{\iota}$, and the size of the spillover effect of ICT capital on TFP, $\gamma$. We estimate $\Omega = (\gamma, \ \sigma_{\iota}^2, \ \rho_{\iota})$ by minimizing,
\begin{eqnarray}\label{equation:min_prob_IRmatching}
\min_{\Omega} \big[  \hat{\Psi} - \Psi(\Omega)  \big]' \Lambda \big[  \hat{\Psi} - \Psi(\Omega)  \big]
\end{eqnarray}
where $\Psi(\Omega)$ denotes the mapping from $\Omega$ to the theoretical impulse responses, and $\hat{\Psi}$ denotes the empirical impulse responses of an ICT shock to TFP, ICTI, consumption, and relative prices shown in Appendix \ref{section:mainSetResults}. $\Lambda$ is a diagonal weighting matrix, which is the inverse of the variance matrix of the empirical impulse responses. These elements serve as weights to impose more precision on some estimates of the optimization problem. Finally, consistent with our choice of horizons presented above, we use a horizon of $40$ periods for each response. % which sum up to a total of $160$ elements for three theoretical parameters only.

@@

Although both the model and impulse-response matching exercise should be considered as preliminary, we believe that they are already able to rationalize the empirical results obtained in Section \ref{section:empirics} and thus convey the main message of this paper. Formally, the question we tried to answer is why a contemporaneous technological shock specific to the ICT sector is able to trigger increases in future total factor productivity even when it is too small to affect TFP today. In other words, why is an ICT-investment shock able to increase the productivity of ICT-good users with lags?

@@

This assumption is legitimate from a mathematical standpoint for two main reasons. First of all, the spillover enters the estimated Solow residual since, being an externality, it is a non-renumerated input. Additionally, since the spillover arises from a state variable, it leads to a dynamics where the benefits of an ICT technology shock show up in the productivity of ICT-capital users with some lags because the accumulation of ICT goods is a slow process. Thus, the spillover effect only comes into play only gradually once enough ICT capital is being accumulated. The slow-moving nature of the state variable capital also means that this effect tends to be persistent, which also helps us match the empirical responses we see in the data.\footnote{We could slow down the dynamics of ICT capital further, for example by assuming investment adjustment costs. Obviously, this would improve the fit of the model since it would allow the externality effect to play out even slower. Our key point here, however, is to show how the spillover is necessary to match the main features of the empirical responses. Without it, the model has no chance to capture the lagged effect of ICT shocks on TFP.}

More importantly, there is important economic intuition underlying our assumption of the spillover. Since the purpose of ICT goods is to enhance communication, the quality and speed of information acquisition and sharing is also determined by the overall diffusion of ICT goods. As a simple example, assume that in 1995 a scholar decides to purchase a computer with an Internet subscription in order to speed up the acquisition of information. She is very satisfied by this investment and does not regret the amount paid. However, at the same time, a former colleague who is now working in a different university decides to buy a computer with an Internet subscription as well. Once she learns of this development, the initial researcher decides to henceforth use her computer to exchange research material with her former colleague as well. They eventually decide to start a new project together, since being in two different universities is no longer a constraint for their joint research. None of the two scholars paid for this additional benefit when they purchased a computer, since they bought it for their personal use and did not expect the other to simultaneously acquire one also. What is actually happening here is that the research production of both researchers has been boosted by the positive externality that also the other colleague is endowed with the same technology. This externality is what we understand under the general-purpose nature of ICT goods and what our model captures with the assumption of the spillover.

@@

In the second part of the paper, we dig deeper into the nature of the link between current ICT investment and future TFP. We rationalize our empirical results by analyzing a two-sector DSGE model \`a la \cite{greenwood1997long} augmented by two main assumptions: (i) the sector with the faster technological improvement produces specifically ICT capital, and (ii) both production functions are augmented with a spillover related to the overall level of ICT capital which is not internalized by the agents. In particular, this last assumption points out our interpretation of the general-purpose nature of ICT capital: the effectiveness of ICT capital is enhanced by its diffusion, and the cost of ICT goods does not reflect this indirect benefit. We use both empirical and theoretical results to estimate main parameters of the model via impulse-response matching to an ICT technology shock. Results show that the parameter which governs the size of the spillover effect is positive and large, suggesting that our interpretation of the general-purpose nature of ICT capital is confirmed by the data. Thus, our investigation suggests not only that ICT is an important driver of medium-run TFP, but specifically that what matters for the evolution of TFP is whether ICT can act as a general-purpose technology. In particular, large externalities from using ICT goods can explain why a sector comprising a tiny fraction of both output and investment can play an important role for future productivity developments.

>>>>

Secondly, a long line of literature going back to the early 1990s postulates that ICT may be a general-purpose technology.\footnote{\cite{basu2003case}, \cite{basu2007information} and \cite{oliner2000resurgence} and are some examples.} % need to include Brynjolfsson's work here, the all-star paper and the Gordon-Mokyr debate.
Following the classical definition of BT,
% Bresnahan Trajtenberg -- need to get citation right
 GPTs are technologies that alter the process of economy-wide production, and thus affect the productivity of the entire economy. This effect can come about  through several channels. For example, investment in ICT goods may trigger accumulation of unobserved complementary capital, as suggested by both \cite{basu2003case} and \cite{basu2007information}. Conversely, as advocated by Brynjolffson, new ICT goods may change managerial practices, thus affecting the production process. We take a broader view in which we simply consider the diffusion of ICT goods to be the crucial element of ICT's general-purpose nature.

To understand what we mean by ICT diffusion, consider again the example of purchasing a mobile phone from the Introduction. Clearly, an individual's productivity improves when acquiring a mobile phone, since she gains access to many functionalities she couldn't enjoy previously. However, the full productivity boost can only materialize when others also use mobile phones. In fact, the more mobile phones there are in the economy, the more productivity gain the individual experiences. Thus, the higher the diffusion of ICT, the higher is its effect on productivity. This is the effect our spillover assumption means to capture. The key exercise, then, of our theoretical model will be to verify via impulse-response matching whether data support the existence of a spillover and thus the hypothesis of ICT as a GPT in the three decades following 1989 in the US economy.

@@

The household in our economy is completely standard. The economy is inhabited by a representative agent who maximizes the expected value of his lifetime utility as given by
\begin{eqnarray}\label{equation:utilFunc}
\E \Bigg[  \sum_{t=0}^{\infty} \beta^t U(c_t,l_t)  \Bigg]
\end{eqnarray}
with 
%\begin{eqnarray*}
%U(c_t,l_t) = \log(c_t) - \frac{\chi}{2} l_t^{2}
%\end{eqnarray*}
\begin{eqnarray*}
U(c_t,l_t) = \log(c_t) - \frac{1}{1+\frac{1}{\chi}} l_t^{1+\frac{1}{\chi}}
\end{eqnarray*}
where $c_t$ and $l_t$ represent consumption and labor at time $t$.

@@

Now we estimate the model's key parameters governing the effect of ICT investment on future TFP by imposing theoretical impulse responses of an ICT shock ($\zeta_t$) to be as close as possible to the empirical ones. Using empirical impulse responses of TFP, ICT investment, consumption, and relative prices, we estimate three parameters: the variance of an ICT technology shock, $\sigma_{\iota}^2$, the persistence of the same shock, $\rho_{\iota}$, and the size of the spillover effect of ICT capital on TFP, $\gamma$. We estimate $\Omega = (\gamma, \ \sigma_{\iota}^2, \ \rho_{\iota})$ by minimizing,
\begin{eqnarray}\label{equation:min_prob_IRmatching}
\min_{\Omega} \big[  \hat{\Psi} - \Psi(\Omega)  \big]' \Lambda \big[  \hat{\Psi} - \Psi(\Omega)  \big]
\end{eqnarray}
where $\Psi(\Omega)$ denotes the mapping from $\Omega$ to the theoretical impulse responses, and $\hat{\Psi}$ denotes the empirical impulse responses of an ICT shock to TFP, ICTI, consumption, and relative prices shown in Appendix \ref{section:mainSetResults}. $\Lambda$ is a diagonal weighting matrix, which is the inverse of the variance matrix of the empirical impulse responses. These elements serve as weights to impose more precision on some estimates of the optimization problem. Finally, consistent with our choice of horizons presented above, we use a horizon of $40$ periods for each response. % which sum up to a total of $160$ elements for three theoretical parameters only.

@@

Although both the model and impulse-response matching exercise are simple and preliminary, we believe that they are already able to rationalize the empirical results obtained in Section \ref{section:empirics} and thus convey the main message of this paper. Formally, the question we tried to answer is why a contemporaneous technological shock specific to the ICT sector is able to trigger increases in future total factor productivity even when it is too small to affect TFP today. In other words, why is an ICT-investment shock able to increase the productivity of ICT-good users with lags?

@@

This assumption is legitimate from a mathematical standpoint for two main reasons. First of all, the spillover enters the estimated Solow residual since, being an externality, it is a non-renumerated input. Additionally, since the spillover arises from a state variable, it leads to a dynamics where the benefits of an ICT technology shock show up in the productivity of ICT-capital users with some lags because the accumulation of ICT goods is a slow process. Thus, the spillover effect only comes into play only gradually once enough ICT capital is being accumulated. The slow-moving nature of the state variable capital also means that this effect tends to be persistent, which also helps us match the empirical responses we see in the data.\footnote{We could slow down the dynamics of ICT capital further, for example by assuming investment adjustment costs. Obviously, this would improve the fit of the model since it would allow the externality effect to play out even slower. Our key point here, however, is to show how the spillover is necessary to match the main features of the empirical responses. Without it, the model has no chance to capture the large lagged effect of ICT shocks on TFP.}

More importantly, there is clear economic intuition underlying our assumption of the spillover. Since the purpose of ICT goods is to enhance communication, the quality and speed of information acquisition and sharing is also determined by the overall diffusion of ICT goods. As a simple example different from the one in the Introduction, assume that in 1995 a scholar decides to purchase a computer with an Internet subscription in order to speed up the acquisition of information. She is very satisfied with this investment and does not regret the amount paid. However, at the same time, a former colleague who is now working in a different university decides to buy a computer with an Internet subscription as well. Once she learns of this development, the initial researcher decides to henceforth use her computer to exchange research material with her former colleague. They eventually decide to start a new project together, since being in two different universities is no longer a constraint for their joint research. None of the two scholars paid for this additional benefit when they purchased a computer, since they bought it for their personal use and did not expect the other to simultaneously acquire one also. What is actually happening here is that the research production of both researchers has been boosted by the positive externality that also the other colleague is endowed with the same technology. This externality is what we understand under the general-purpose nature of ICT goods and what our model captures with the assumption of the spillover.

@@

In the second part of the paper, we dig deeper into the nature of the link between current ICT investment and future TFP. We rationalize our empirical results by analyzing a two-sector DSGE model \`a la \cite{greenwood1997long} augmented by two main assumptions: (i) the sector with the faster technological improvement produces specifically ICT capital, and (ii) both production functions are augmented with a spillover related to the overall level of ICT capital which is not internalized by the agents. In particular, this last assumption points out our interpretation of the general-purpose nature of ICT capital: the effectiveness of ICT capital is enhanced by its diffusion, and the cost of ICT goods does not reflect this indirect benefit.

We use both empirical and theoretical results to estimate main parameters of the model via impulse-response matching to an ICT technology shock. Results show that the parameter which governs the size of the spillover effect is positive and large, suggesting that our interpretation of the general-purpose nature of ICT capital is confirmed by the data. Thus, our investigation suggests not only that ICT is an important driver of medium-run TFP, but specifically that what matters for the evolution of TFP is whether ICT can act as a general-purpose technology. In particular, large externalities from using ICT goods can explain why a sector comprising a tiny fraction of both output and investment can play an important role for future productivity developments.
